

```{r}

#remove scientific notation
options(scipen=999)

### Run this line to install all packages, if needed. 
#install.packages(c("sf", "tidyverse", "leaflet", "mapview", "janitor", "lubridate", "tidycensus", "lwgeom", "tigris"))

### Call our packages into our script here.
library(sf)
library(tidyverse)
library(janitor)
library(mapview)
library(lubridate)
library(tidycensus)
library(tigris)
library(lwgeom)

sf_use_s2(FALSE)


```


## Importing, inspecting and visualizing geo data

Helpful Indianapolis codes:
- Indiana is state fips 18
- Indianapolis is in Marion County, which is county fips 18097.

Where to download/find these GIS files for your state!
- **Schools:** [The NCES EDGE open data portal](https://data-nces.opendata.arcgis.com/) contains all sorts of helpful data, including school locations.
- **Neighborhoods:** check your city GIS site
- **Toxic Release Inventory:** Multiple years and states [available here](https://www.epa.gov/toxics-release-inventory-tri-program/tri-basic-data-files-calendar-years-1987-present)
- **Census Tigerline URLs:** [pick your vintage here](https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html), then click FTP Archive and navigate to the shapefile you need.need.

```{r}

#importing shapefiles and geojson
schools <- st_read("../GIS/nces-in-schools.geojson") |>
  # a function that cleans up column names
  clean_names()

#importing from zip
hoods = st_read("../GIS/Indy_Neighborhoods/Indy_Neighborhoods.shp") |>
  clean_names()

#importing from library/api
tracts <- tracts(state = "IN", cb = FALSE) |>
  clean_names()

#sf::write_sf(tracts, "../GIS/tracts.geojson")

#importing from csv

tri <- read_csv("../GIS/epa-tri-2024-in.csv") |>
  clean_names() |>
  st_as_sf(coords = c("x13_longitude", "x12_latitude"), crs = 4326)

# we are going to remove those pesky numbers in our column names because they are annoying
names(tri) <- str_remove(names(tri), "^x\\d+_")

```

```{r}

#exploring geospatial data is just like exploring any other dataframe

glimpse(schools)

```

```{r}

head(schools)

```

```{r}

#lets take a closer look at the geometry column

schools |>
  select(geometry)

```

```{r}

# we can also look at the projection of the data using mapview

st_crs(schools)

```

```{r}

#now you try looking at the other geodataframes crs

```


Cam is going to talk about projections here for a minute. They're important friends.

```{r}

mapview(schools)

```

## Finding hidden ARC data
Have you ever come across some really great geo data on a city's website but there's no download button? If it's an ARCGIS site you might be able to access the data by going through the backend!

Let's demo that now. If you're coming back to this later, here's a video of how to find and download ARCGIS data: https://www.youtube.com/watch?v=bqSBfOTTGwU

```{r}

#let's load the crime data we just downloaded from the Indy PD's ARCGIS
#https://gis.indy.gov/server/rest/services/IMPD
#https://gis.indy.gov/server/rest/services/IMPD/IMPD_Public_Data/FeatureServer/1/query?where=1%3D1&objectIds=&time=&geometry=&geometryType=esriGeometryPoint&inSR=&defaultSR=&spatialRel=esriSpatialRelIntersects&distance=&units=esriSRUnit_Foot&relationParam=&outFields=*&returnGeometry=true&maxAllowableOffset=10000&geometryPrecision=&outSR=&havingClause=&gdbVersion=&historicMoment=&returnDistinctValues=false&returnIdsOnly=false&returnCountOnly=false&returnExtentOnly=false&orderByFields=&groupByFieldsForStatistics=&outStatistics=&returnZ=false&returnM=false&multipatchOption=xyFootprint&resultOffset=&resultRecordCount=&returnTrueCurves=false&returnExceededLimitFeatures=false&quantizationParameters=&returnCentroid=false&timeReferenceUnknownClient=false&maxRecordCountFactor=&sqlFormat=none&resultType=&featureEncoding=esriDefault&datumTransformation=&cacheHint=false&f=geojson


crime <- read_sf("../GIS/indy-nibrs.geojson") |>
  clean_names()

glimpse(crime)

mapview(crime)

```

Now we're going to load our Census API key so we can access some Census data!

We are going to use TidyCensus to access the Census API. TidyCensus is a wrapper around the Census API that makes it easier to use in R. 

```{r}

#sometimes an API requires an access key
#here's an example oh how to get Census ACS 5-yr data via the API
#and more info on the various APIs https://www.census.gov/data/developers/data-sets/acs-5year.html
#NOTE: different tables (detailed, subject, etc) have different geography levels available



data_2024 <- get_acs(geography = "tract", 
        variables = c("B01001_001", "B11016_001", "B19013_001"), #picking the variables we want
        state = "IN", #setting the state to minnesota
        geometry = FALSE, # not including the geometry
        cache_table = TRUE) |> #caching the data so we don't call the API multiple times
  select(-moe) |> #removing the margin of error columns
  pivot_wider(values_from = estimate, names_from = variable) |> #pivoting the data wider so each column is a variable
  rename(total_pop = B01001_001, #renaming the columns to be more descriptive
         households = B11016_001, 
         median_income = B19013_001) |>
  clean_names()


#use this is tidyverse doesn't work
#data_2024 <- read_csv("../data/analyzed/2024-ACS5yr-indy-in-tract-data.csv")

data_2024 |>
  glimpse()

```

## Spatial joins and filtering

```{r error=TRUE}

crime_in_hoods <- st_join(crime, hoods, join = st_within)

```

Remember, the cordination system of the two datasets must match for the join to work. Let's fix that.

```{r}

crime <- crime |>
  st_transform(crs=4326) |>
  st_make_valid()

hoods <- hoods |>
  st_transform(crs=4326) |>
  st_make_valid()

```

Now let's try again.

```{r}

#in r, we must specify left = FALSE to ensure only the potholes in the neighborhoods are returned
crime_in_hoods <- st_join(potholes, hoods, join = st_within, left = FALSE)

glimpse(crime_in_hoods)
mapview(crime_in_hoods) +
  mapview(hoods)

```

## Count points in polygon

My fav way to do a point in polygon is to combine a spatial join and a group_by. First, the spatial join lets us connect a neighborhood name with each pothole, next, the group_by will let us count the potholes present in each neighborhood. 

```{r}

crime_in_hoods |>
  group_by(name) |>
  summarize(crime_count = n()) |>
  arrange(desc(crime_count))

```

```{r}

#if we feel like knowing more about the types of crimes in each hood, we can do a pivot_table instead

crime_in_hoods |>
  st_drop_geometry() |> #this time we will drop geometry so it doesn't show up in our aggregate table
  group_by(name, cr_desc) |>
  summarize(crime_count = n()) |>
  pivot_wider(names_from = cr_desc, values_from = crime_count, values_fill = 0) |> #pivoting wider
  clean_names() |>
  mutate(total = rowSums(across(where(is.numeric)))) |> #adding a total column
  arrange(desc(total)) |> #sorting by total descending
  relocate(total, .after = name) #moving the total column next to the neighborhood name

```

## Crosswalks

Generally, demographic data like population and median income isn't calculated for custom geographies like neighborhoods. But no matter! We can calculate that sort of information for ourselves using crosswalks.

In this example, we are going to cross Census tracts with Indy neighborhoods.

Cam is going to take a minute and talk about why, in the wild, we might want to use block groups instead of tracts. Note that some of the Census data tables are only available at the tract level.

```{r}

#join
tracts_data <- tracts |>
  left_join(data_2024, join_by("geoid"))

#we need to make sure both gdfs are in the same projection, but unlike the spatial join
#we need to make sure that projection uses meters or feet because we'll be calculating area
#this projection is specific to the Twin Cities area: https://epsg.io/6505

tracts_data <- tracts_data |>
  st_transform(crs = 6505)

hoods <- hoods |>
  st_transform(crs = 6505)

#and we'll need to calculate the area of the tracts before and after we do the intersection so we can
#calculate the actual estimated count of people of different races in each tract segmemt.

tracts_data <- tracts_data |>
  mutate(pre_area = st_area(geometry))

#now let's do that intersection!

#Ryan here: For some reason, this code returns an error. I cannot figure out why,. It gives quite possibly the least descriptive error I have ever seen in my life. St_join should work the same as calling the intersection function directly, but tis not in this case. Very annoying.
#tracts_hoods <-  st_join(tracts_data, hoods, join = st_intersection)

# So we will jsut do it this way instead
tracts_hoods <- st_intersection(tracts_data, hoods)

#and calculate the area post intersection.
tracts_hoods <- tracts_hoods |>
  mutate(post_area = st_area(geometry))

#from the pre and post area calculations we create a percentage
tracts_hoods <- tracts_hoods |>
  mutate(pct_area = post_area/pre_area)

#and then we multiply our tract data counts by that percentage
#to get estimates per segments of the tract that falls within each hood

tracts_hoods <- tracts_hoods |>
  mutate(
    post_pop = total_pop * pct_area,
    post_households = households * pct_area,
    post_income = median_income * pct_area
  )

#groupby the neighborhood name so we can join back to our neighborhood shapes

hoods_data = tracts_hoods |>
  group_by(name) |>
  summarise(
    post_pop = sum(post_pop),
    post_households = sum(post_households),
    post_icome = mean(post_households)
  )

hoods_data |>
  head()

```

## Buffers
Now, let's work with our schools and toxic release inventory data! I think it would be interesting to know which schools have a facility that releases toxic air or water chemicals near it. Shall we say 1 mile? If this were for an actual story, we would talk to an expert in the field to figure out the best distance to use.

Cam is gonna talk about finding experts to help inform your geographic analysis.

```{r}

#first, we have to change the projection to get it into feet

schools <- schools |>
  st_transform(crs = 2965)

# Next, we draw a 1 mile buffer around each of our schools. The buffer distance depends on 
# the units of the CRS. Since our CRS is in feet, we need to convert 1 mile to feet.
schools_buffers <- schools |> 
  st_buffer(5280)


# keep only the columns we need
schools_buffers <- schools_buffers |>
  select(ncessch, name, city, zip, nmcnty, nmcbsa)

mapview(schools_buffers)

```

```{r}

#we're going to group the TRI by facility because with this dataset, a single facility will have 
#many rows, one for each chemical released

unique_tri <- tri |>
  group_by(trifd, facility_name, street_address, city, county) |>
  summarise(chemicals_released = n())

# us R users never lose the geometry column when we group_by, but droppping it could speed things up on big files
unique_tri |>
  arrange(desc(chemicals_released)) |>
  head()

```
```{r}

mapview(tri)

```


```{r}

school_buffers <- schools_buffers |>
  st_transform(4326)

unique_tri <- unique_tri |>
  st_transform(4326)

tri_schools <- st_join(unique_tri, school_buffers, join = st_within)

print(nrow(tri))

print(nrow(tri_schools |>
             unique()))

tri_schools |>
  head()
```

If you're asking yourself "why that CRS?" it's because when you do a spatial join with points, you need to everything to be in a geometric CRS rather than a projected CRS. I don't know man, ask Cam. She's way smarter than I am.

If you're further asking yourself "Why are there some schools in there more than once!?". It's because some schools have more than one facility within one mile of them. Le YIPE! 

But we know how to deal with that, right? We can groupby the school and count the facilities!

```{r}

#if you're wondering why there are MORE tri_schools records than tri records, it's because
#some facilities are within a mile of more than one school. Le YIPE!
#but we know how to deal with that, right? We can groupby the school and count the facilities

tri_per_school <- tri_schools |>
  group_by(name, city.x, nmcnty) |>
  summarise(tri_facilities = n())

tri_per_school |>
  arrange(desc(tri_facilities)) |>
  head()

```

## Calculating distance
So now we know which schools are within a mile of a TRI facility... but what if I wanted to know about the school with the closest TRI facility? Or maybe calculate all of the distances between schools and their closes TRI facility?

```{r}

# you're gonna hate me... but we need to change these damn projections back to 6505 
# cause we need the feet!

schools <- schools |>
  st_transform(6505)

unique_tri <- unique_tri |>
  st_transform(6505)

#In R, we first have to build a list of features that are closest to each school
nearest_indices <- st_nearest_feature(schools, unique_tri)

# Because it is put out in the order of the dataframe, we can just cbind it to the existing data
closest_tri_school <- schools %>%
  mutate(nearest_tri_geom = st_geometry(unique_tri)[nearest_indices]) %>%
  #then calculate the distance between them
  mutate(distance_m = st_distance(geometry, nearest_tri_geom, by_element = TRUE))

glimpse(closest_tri_school)

```

##Exporting for your friends

```{r}
library(htmlwidgets)

#writing mapview output to an object
map <- mapview(closest_tri_school)

#saving the object to a html file
saveWidget(map@map, file = "closest_tri_school.html", selfcontained = TRUE)

```


